---
title: "Lab 05 - wrangling spacial data"
author: Ben Waggener
date: 3/21/2025
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```



### Exercise 1

"Filter the Denny’s data frame for Alaska (AK) and save the result as dn_ak. How many Denny’s locations are there in Alaska?"
```{r ex1_dns_alaska}
dn_ak <- dennys %>%
  filter(state == "AK")
nrow(dn_ak)

```
"Now, do the same for La Quinta data frame for Alaska (AK) and save the result as lq_ak. How many La Quinta locations are there in Alaska?"
```{r ex1_lq_alaska}
lq_ak <- laquinta %>%
  filter(state == "AK")
nrow(lq_ak)

```
There are 3 dennys in Alaska and 2 La Quintas in Alaska.

### Exercise 2
"Next we will be calculating the distance between all Denny’s and all La Quinta locations in Alaska."

there are 6 possible pairings between the 3 Denny's and 2 La Quintas in Alaska. 

### Exercise 3
"Let’s join the data on Denny’s and La Quinta locations in Alaska, and take a look at what it looks like:"

```{r ex3_fulljoin}
dn_lq_ak <- full_join(dn_ak, lq_ak, 
                      by = "state")
dn_lq_ak
```


### Exercise 4
"How many observations are in the joined dn_lq_ak data frame? What are the names of the variables in this data frame."

variable names
```{r ex4_nrow_ncol}
nrow(dn_lq_ak)
ncol(dn_lq_ak)
names(dn_lq_ak)
```
There are 6 observations and 11 variables in the joined data frame. The variables are "address.x"   "city.x"      "state"       "zip.x"       "longitude.x" "latitude.x"  "address.y"   "city.y"     
  "zip.y"       "longitude.y" "latitude.y" 
...

### Exercise 5

...

### Exercise 6

...

Add exercise headings as needed.
